The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles

### 2022-11-09 ms/ms.Rmd
<edit ms>

---
title: |
  The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles 
type: ORIGINAL RESEARCH ARTICLE
author:
  - name: R. Alan Covey
    affil: a
    email: R.Alan.Covey@austin.utexas.edu
  - name: Robert Z. Selden, Jr.
    affil: b
    email: zselden@sfasu.edu
  - name: Astrid Runggaldier
    affil: c
    email: astrid@austin.utexas.edu
  - name: Nicole D. Payntar
    affil: a
    email: npayntar@utexas.edu
affiliation:
  - num: a
    address: |
      Department of Anthropology, The University of Texas at Austin
  - num: b
    address: |
      Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University 
  - num: c
    address: |
      Department of Art and Art History, The University of Texas at Austin; The Mesoamerica Center, The University of Texas at Austin
  - num: d
    address: |
      Department of Anthropology, The University of Texas at Austin
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  After centuries of looting along Peru’s North Coast, archaeologists acknowledge that the pottery of the Chimú Empire is one of the most collected Andean artifacts, but also one of the most poorly understood.  Much of the enduring classificatory uncertainty comes from the problematic provenance of most Chimú vessels, and the fact that the distinctive blackware identified as Chimú represents the production of workshops from across an extensive area, during periods of regional political decentralization (CE 900-1200), imperial growth (CE 1200-1450), and foreign conquest by the Inca (CE 1450-1535) and Spanish (after CE 1532) empires.  This chapter builds on previous seriations and field observations, using geometric morphometric analysis of a sample of 3D-scanned Chimú bottles from publicly held collections at the University of Texas at Austin.  Since Chimú bottles were formed in workshops using molds, variations in vessel shape can serve as indicators of variable practices.  We compare a sample of common Chimú blackware bottles with a sample of _Inca-Chimú_ vessels that carry features typical of the short Inca occupation of the North Coast.  Differences between the two vessel forms offer new lines of evidence that can guide more precise classifications, as well as new interpretations of the imperial history of Peru’s North Coast.
keywords: |
  Sections; lists; figures; tables; mathematics; fonts; references; appendices
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

The history surrounding the collection of Chimú blackwares helps to explain why this ubiquitous artifact class remains poorly understood by professional archaeologists.  When viewed as a byproduct of other collecting activities that unfolded over almost a half millennium, the Chimú ceramic corpus can be seen for what it is: an ambiguous assemblage of mostly orphaned objects.  Almost all Chimú vessels lack a detailed archaeological provenience, and the chain of ownership for most museum pieces can rarely be traced back to Peru.  Because most “Chimú” vessels cannot be affiliated to a particular context in coastal Peru, their identification derives from their stylistic attributes, even though some of the earliest Chimú ceramics documented in archaeological excavations appeared at the Inca site of Machu Picchu (Bingham 1915:269) and the coastal creation shrine of Pachacamac, two sites that lie well beyond the political control of the Chimú Empire (Uhle 1902:756).

This contextual ambiguity intersects with a classificatory problem for Andean archaeology.  Many treat Chimú pottery as typical of the Late Intermediate Period (c. 900-1476) on the north coast of Peru, a huge and undifferentiated period during which the Chimú polity established its capital and local dominion in the Moche Valley, carried out multiple waves of imperial expansion to the north and south, and fell under the hegemony of Inca imperial rule (1476-1532).  Excavators have noted that similar blackwares continued to be produced during the half century of Inca rule, and into the earliest years of the Spanish colonial period.  Without knowing when and where a particular museum object comes from, archaeologists have faced difficulties in subdividing Chimú pottery into phases and substyles, which would enable researchers to study changes and variations within the overall assemblage.

@@

From early on, classifications of Chimú ceramics focused on vessel attributes.  Beuchat (1918:652-654) noted that coastal tombs had yielded “infinite” formal variations of blackware pots now found in museum collections.  He placed these into five principal categories: (1) globular vessels with rectangular fields of low-relief decoration; (2) geometric vessels (cuboid, ellipsoids, ovular); (3) phytomorphic vessels shaped like squash or fruits; (4) zoomorphic vessels representing a broad range of fauna; and (5) anthropomorphic vessels.  This classification emphasized aesthetic characteristics of vessel shape, rather than the kinds of formal and functional differences seen in later typologies, and Beuchat acknowledged the diversity found within each of his categories, especially the phytomorphic, zoomorphic, and anthropomorphic vessels.  Without any provenience information to guide his analysis, Beuchat did not attempt to seriate Chimú pottery or to distinguish regional variations.

In 1926, Alfred Kroeber made the first concerted effort to describe different phases of Chimú ceramic production, based on a brief visit to the Trujillo area that he made as part of a Field Museum expedition to the central coast.  Kroeber noted previous efforts by Uhle (1902) to differentiate between early and late ceramics from Trujillo, which could be tied to the Moche site of Huaca de la Luna/Huaca del Sol and the Chimú capital of Chan Chan.  Kroeber used the term _Late Chimú_ to refer to Trujillo area blackwares that sometimes were found with Inca ceramics, a designation that encompasses Late Intermediate Period (c. 900-1400s), Inca (1400s-1530s) and some early Colonial pottery.  Based on provenance information that recorded the district where some Late Chimú vessels originated, Kroeber (1926:11) identified a broad distribution that stretched from Piura to Casma, with some vessels found as far south as the Nasca drainage.  To build a typology and sequence of Late Chimú vessels, he studied the pottery from several collections that he purchased in Trujillo on behalf of the Field Museum.

Kroeber’s shape typology for Late Chimú pottery discusses vessels that were common in Moche and Chimú assemblages, and he lists (1926:22) about a dozen Late Chimú shapes, including stirrup-mouth vessels, double jars, double-spout vessels, head-and-spout vessels, figure-and-spout vessels, globular bowls, lipped pots (with and without handles), jars with flaring mouth, flat-handled jars, and jars with tapering spouts.  The list of shapes also includes Inca narrow-mouth jars (_aryballus_) and cups (_kero_) in the Late Chimú assemblage.  In his decorative analysis of vessels in private collections, Kroeber (27-28) identified the _face vase_ and _rotund figure jar_ as additional shapes found in the assemblage.  The discussion of Late Chimú blackware recorded the relative abundance of the most common shapes found in two private collections that he saw while visiting Trujillo.  The Mansiche and Jacoms collections comprised more than 200 vessels, but Kroeber estimated that hundreds of utilitarian pots, plates, and jars had been _rejected_ by the collectors when they purchased vessels from the huaqueros who excavated them near Chan Chan.

In his study of collected ceramics from Trujillo, Kroeber established the outline of a shape typology for Chimú blackware vessels, although it was not a comprehensive representation of ceramics in use on the north coast during the periods of Chimú, Inca, and Spanish imperial expansion.  Building on the earlier observations of the high degree of diversity seen in Chimú blackwares, Kroeber made some important observations that inform subsequent studies.  First, this pottery is widely distributed along the north coast of Peru, although Kroeber (1930:97) argued for strong continuity between regions, which would suggest that regional variations account for little of the overall diversity in the Chimú assemblage.  Second, he noted that blackwares were strongly associated with mortuary contexts, and such pottery was not commonly seen on the surface of the urban core of Chan Chan.  Such vessels were thus more indicative of Chimú funerary practices than everyday life.  Finally, Kroeber described how the private collections that local dealers assembled and sold to museums were already curated in a way that eliminated certain vessel shapes, so that museum collections could be considered representative of the original mortuary assemblage.  As Kroeber noted (1930:95) in a later visit to the north coast, one Trujillo collector “refused to purchase from the huaqueros most of the unhandled jars, exceptions being made in favor of effigy pieces, or occasional plain ones when several vessels were bought in a lot in order to acquire one or two attractive ones.”

@@

In the years following Kroeber’s work, analysis of Chimú ceramics continued to be oriented around the integration of regional Andean sequences, and the pottery of the Chimú Empire continued to be discussed alongside Moche ceramics that were considered to be Early Chimú (e.g., Bennett 1937; Willey 1945).  It was not until the 1960s that a concerted effort was made to move beyond the master sequence, to developed a more fine-grained description of the assemblage associated with the Chimú Empire as it developed, expanded, and fell under Inca and Spanish dominion.  In 1966, Harry Scheele and Thomas Patterson published a “preliminary seriation” of Chimú ceramics.  They noted (1966:15) that one reason that so little work had been done to classify such material was its abundance: “probably no other pottery style from the Americas is so well represented on the shelves of museums and private collectors throughout the world.”

>>>>

---
title: |
  The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles 
type: ORIGINAL RESEARCH ARTICLE
author:
  - name: R. Alan Covey
    affil: a
    email: R.Alan.Covey@austin.utexas.edu
  - name: Robert Z. Selden, Jr.
    affil: b
    email: zselden@sfasu.edu
  - name: Astrid Runggaldier
    affil: c
    email: astrid@austin.utexas.edu
  - name: Nicole D. Payntar
    affil: a
    email: npayntar@utexas.edu
affiliation:
  - num: a
    address: |
      Department of Anthropology, The University of Texas at Austin
  - num: b
    address: |
      Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University 
  - num: c
    address: |
      Department of Art and Art History, The University of Texas at Austin; The Mesoamerica Center, The University of Texas at Austin
  - num: d
    address: |
      Department of Anthropology, The University of Texas at Austin
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  After centuries of looting along Peru’s North Coast, archaeologists acknowledge that the pottery of the Chimú Empire is one of the most collected Andean artifacts, but also one of the most poorly understood.  Much of the enduring classificatory uncertainty comes from the problematic provenance of most Chimú vessels, and the fact that the distinctive blackware identified as Chimú represents the production of workshops from across an extensive area, during periods of regional political decentralization (CE 900-1200), imperial growth (CE 1200-1450), and foreign conquest by the Inca (CE 1450-1535) and Spanish (after CE 1532) empires.  This chapter builds on previous seriations and field observations, using geometric morphometric analysis of a sample of 3D-scanned Chimú bottles from publicly held collections at the University of Texas at Austin.  Since Chimú bottles were formed in workshops using molds, variations in vessel shape can serve as indicators of variable practices.  We compare a sample of common Chimú blackware bottles with a sample of _Inca-Chimú_ vessels that carry features typical of the short Inca occupation of the North Coast.  Differences between the two vessel forms offer new lines of evidence that can guide new interpretations of the imperial history of Peru’s North Coast.
keywords: |
  Computational Archaeology; Museum Studies; Digital Humanities; Non-Western Art History; STEM; STEAM
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

The history surrounding the collection of Chimú blackwares helps to explain why this ubiquitous artifact class remains poorly understood by professional archaeologists.  When viewed as a byproduct of collecting activities that unfolded over almost a half millennium, the Chimú ceramic corpus can be seen for what it is: an ambiguous assemblage of mostly orphaned objects.  Almost all Chimú vessels lack a detailed archaeological provenience, and the chain of ownership for most museum pieces can rarely be traced back to Peru.  Because most “Chimú” vessels cannot be affiliated to a particular context in coastal Peru, their identification derives from their stylistic attributes, even though some of the earliest Chimú ceramics documented in archaeological excavations appeared at the Inca site of Machu Picchu \citep[269]{RN7063} and the coastal creation shrine of Pachacamac, two sites that lie well beyond the political control of the Chimú Empire \citep[756]{RN11154}.

This contextual ambiguity intersects with a classificatory problem for Andean archaeology.  Many treat Chimú pottery as typical of the Late Intermediate Period (CE 900-1476) on the north coast of Peru, a huge and undifferentiated period during which the Chimú polity established its capital and local dominion in the Moche Valley, carried out multiple waves of imperial expansion to the north and south, and fell under the hegemony of Inca imperial rule (CE 1476-1532).  Excavators have noted that similar blackwares continued to be produced during the half century of Inca rule, and into the earliest years of the Spanish colonial period.  Without knowing when and where a particular museum object comes from, archaeologists have faced difficulties in subdividing Chimú pottery into phases and substyles, which would enable researchers to study changes and variations within the overall assemblage.

@@

From early on, classifications of Chimú ceramics focused on vessel attributes. \citet[652-654]{RN11152} noted that coastal tombs had yielded “infinite” formal variations of blackware pots now found in museum collections.  He placed these into five principal categories: (1) globular vessels with rectangular fields of low-relief decoration; (2) geometric vessels (cuboid, ellipsoids, ovular); (3) phytomorphic vessels shaped like squash or fruits; (4) zoomorphic vessels representing a broad range of fauna; and (5) anthropomorphic vessels.  This classification emphasized aesthetic characteristics of vessel shape, rather than the kinds of formal and functional differences seen in later typologies, and Beuchat acknowledged the diversity found within each of his categories, especially the phytomorphic, zoomorphic, and anthropomorphic vessels.  Without any provenience information to guide his analysis, Beuchat did not attempt to seriate Chimú pottery or to distinguish regional variations.

In 1926, Alfred Kroeber made the first concerted effort to describe different phases of Chimú ceramic production, based on a brief visit to the Trujillo area that he made as part of a Field Museum expedition to the central coast.  Kroeber noted previous efforts by \citet{RN11154} to differentiate between early and late ceramics from Trujillo, which could be tied to the Moche site of Huaca de la Luna/Huaca del Sol and the Chimú capital of Chan Chan.  Kroeber used the term _Late Chimú_ to refer to Trujillo area blackwares that sometimes were found with Inca ceramics, a designation that encompasses Late Intermediate Period (CE 900-1400s), Inca (CE 1400s-1530s) and some early Colonial pottery.  Based on provenance information that recorded the district where some Late Chimú vessels originated, \citet[11]{RN11151} identified a broad distribution that stretched from Piura to Casma, with some vessels found as far south as the Nasca drainage.  To build a typology and sequence of Late Chimú vessels, he studied the pottery from several collections that he purchased in Trujillo on behalf of the Field Museum.

Kroeber’s shape typology for Late Chimú pottery discusses vessels that were common in Moche and Chimú assemblages, and he lists \citep[22]{RN11151} about a dozen Late Chimú shapes, including stirrup-mouth vessels, double jars, double-spout vessels, head-and-spout vessels, figure-and-spout vessels, globular bowls, lipped pots (with and without handles), jars with flaring mouth, flat-handled jars, and jars with tapering spouts.  The list of shapes also includes Inca narrow-mouth jars (_aryballus_) and cups (_kero_) in the Late Chimú assemblage.  In his decorative analysis of vessels in private collections, \citet[27-28]{RN11151} identified the _face vase_ and _rotund figure jar_ as additional shapes found in the assemblage.  The discussion of Late Chimú blackware recorded the relative abundance of the most common shapes found in two private collections that he saw while visiting Trujillo.  The Mansiche and Jacoms collections comprised more than 200 vessels, but Kroeber estimated that hundreds of utilitarian pots, plates, and jars had been _rejected_ by the collectors when they purchased vessels from the huaqueros who excavated them near Chan Chan.

In his study of collected ceramics from Trujillo, Kroeber established the outline of a shape typology for Chimú blackware vessels, although it was not a comprehensive representation of ceramics in use on the north coast during the periods of Chimú, Inca, and Spanish imperial expansion.  Building on the earlier observations of the high degree of diversity seen in Chimú blackwares, Kroeber made some important observations that inform subsequent studies.  First, this pottery is widely distributed along the north coast of Peru, although \citet[97]{RN11155} argued for strong continuity between regions, which would suggest that regional variations account for little of the overall diversity in the Chimú assemblage.  Second, he noted that blackwares were strongly associated with mortuary contexts, and such pottery was not commonly seen on the surface of the urban core of Chan Chan.  Such vessels were thus more indicative of Chimú funerary practices than everyday life.  Finally, Kroeber described how the private collections that local dealers assembled and sold to museums were already curated in a way that eliminated certain vessel shapes, so that museum collections could be considered representative of the original mortuary assemblage.  As Kroeber noted \citep[95]{RN11155} in a later visit to the north coast, one Trujillo collector “refused to purchase from the huaqueros most of the unhandled jars, exceptions being made in favor of effigy pieces, or occasional plain ones when several vessels were bought in a lot in order to acquire one or two attractive ones.”

@@

In the years following Kroeber’s work, analysis of Chimú ceramics continued to be oriented around the integration of regional Andean sequences, and the pottery of the Chimú Empire continued to be discussed alongside Moche ceramics that were considered to be Early Chimú \citep{RN11156,RN11157}.  It was not until the 1960s that a concerted effort was made to move beyond the master sequence, to developed a more fine-grained description of the assemblage associated with the Chimú Empire as it developed, expanded, and fell under Inca and Spanish dominion.  In 1966, Harry Scheele and Thomas Patterson published a “preliminary seriation” of Chimú ceramics.  They noted \citep[15]{RN11158} that one reason that so little work had been done to classify such material was its abundance: “probably no other pottery style from the Americas is so well represented on the shelves of museums and private collectors throughout the world.”

### 2022-11-09 ms/ms.tex
<edit ms>

\begin{abstract}
After centuries of looting along Peru's North Coast, archaeologists
acknowledge that the pottery of the Chimú Empire is one of the most
collected Andean artifacts, but also one of the most poorly understood.
Much of the enduring classificatory uncertainty comes from the
problematic provenance of most Chimú vessels, and the fact that the
distinctive blackware identified as Chimú represents the production of
workshops from across an extensive area, during periods of regional
political decentralization (CE 900-1200), imperial growth (CE
1200-1450), and foreign conquest by the Inca (CE 1450-1535) and Spanish
(after CE 1532) empires. This chapter builds on previous seriations and
field observations, using geometric morphometric analysis of a sample of
3D-scanned Chimú bottles from publicly held collections at the
University of Texas at Austin. Since Chimú bottles were formed in
workshops using molds, variations in vessel shape can serve as
indicators of variable practices. We compare a sample of common Chimú
blackware bottles with a sample of \emph{Inca-Chimú} vessels that carry
features typical of the short Inca occupation of the North Coast.
Differences between the two vessel forms offer new lines of evidence
that can guide more precise classifications, as well as new
interpretations of the imperial history of Peru's North Coast.
\end{abstract}

\begin{keywords}
Sections; lists; figures; tables; mathematics; fonts; references;
appendices
\end{keywords}

@@

The history surrounding the collection of Chimú blackwares helps to
explain why this ubiquitous artifact class remains poorly understood by
professional archaeologists. When viewed as a byproduct of other
collecting activities that unfolded over almost a half millennium, the
Chimú ceramic corpus can be seen for what it is: an ambiguous assemblage
of mostly orphaned objects. Almost all Chimú vessels lack a detailed
archaeological provenience, and the chain of ownership for most museum
pieces can rarely be traced back to Peru. Because most ``Chimú'' vessels
cannot be affiliated to a particular context in coastal Peru, their
identification derives from their stylistic attributes, even though some
of the earliest Chimú ceramics documented in archaeological excavations
appeared at the Inca site of Machu Picchu (Bingham 1915:269) and the
coastal creation shrine of Pachacamac, two sites that lie well beyond
the political control of the Chimú Empire (Uhle 1902:756).

This contextual ambiguity intersects with a classificatory problem for
Andean archaeology. Many treat Chimú pottery as typical of the Late
Intermediate Period (c.~900-1476) on the north coast of Peru, a huge and
undifferentiated period during which the Chimú polity established its
capital and local dominion in the Moche Valley, carried out multiple
waves of imperial expansion to the north and south, and fell under the
hegemony of Inca imperial rule (1476-1532). Excavators have noted that
similar blackwares continued to be produced during the half century of
Inca rule, and into the earliest years of the Spanish colonial period.
Without knowing when and where a particular museum object comes from,
archaeologists have faced difficulties in subdividing Chimú pottery into
phases and substyles, which would enable researchers to study changes
and variations within the overall assemblage.

@@

From early on, classifications of Chimú ceramics focused on vessel
attributes. Beuchat (1918:652-654) noted that coastal tombs had yielded
``infinite'' formal variations of blackware pots now found in museum
collections. He placed these into five principal categories: (1)
globular vessels with rectangular fields of low-relief decoration; (2)
geometric vessels (cuboid, ellipsoids, ovular); (3) phytomorphic vessels
shaped like squash or fruits; (4) zoomorphic vessels representing a
broad range of fauna; and (5) anthropomorphic vessels. This
classification emphasized aesthetic characteristics of vessel shape,
rather than the kinds of formal and functional differences seen in later
typologies, and Beuchat acknowledged the diversity found within each of
his categories, especially the phytomorphic, zoomorphic, and
anthropomorphic vessels. Without any provenience information to guide
his analysis, Beuchat did not attempt to seriate Chimú pottery or to
distinguish regional variations.

In 1926, Alfred Kroeber made the first concerted effort to describe
different phases of Chimú ceramic production, based on a brief visit to
the Trujillo area that he made as part of a Field Museum expedition to
the central coast. Kroeber noted previous efforts by Uhle (1902) to
differentiate between early and late ceramics from Trujillo, which could
be tied to the Moche site of Huaca de la Luna/Huaca del Sol and the
Chimú capital of Chan Chan. Kroeber used the term \emph{Late Chimú} to
refer to Trujillo area blackwares that sometimes were found with Inca
ceramics, a designation that encompasses Late Intermediate Period
(c.~900-1400s), Inca (1400s-1530s) and some early Colonial pottery.
Based on provenance information that recorded the district where some
Late Chimú vessels originated, Kroeber (1926:11) identified a broad
distribution that stretched from Piura to Casma, with some vessels found
as far south as the Nasca drainage. To build a typology and sequence of
Late Chimú vessels, he studied the pottery from several collections that
he purchased in Trujillo on behalf of the Field Museum.

Kroeber's shape typology for Late Chimú pottery discusses vessels that
were common in Moche and Chimú assemblages, and he lists (1926:22) about
a dozen Late Chimú shapes, including stirrup-mouth vessels, double jars,
double-spout vessels, head-and-spout vessels, figure-and-spout vessels,
globular bowls, lipped pots (with and without handles), jars with
flaring mouth, flat-handled jars, and jars with tapering spouts. The
list of shapes also includes Inca narrow-mouth jars (\emph{aryballus})
and cups (\emph{kero}) in the Late Chimú assemblage. In his decorative
analysis of vessels in private collections, Kroeber (27-28) identified
the \emph{face vase} and \emph{rotund figure jar} as additional shapes
found in the assemblage. The discussion of Late Chimú blackware recorded
the relative abundance of the most common shapes found in two private
collections that he saw while visiting Trujillo. The Mansiche and Jacoms
collections comprised more than 200 vessels, but Kroeber estimated that
hundreds of utilitarian pots, plates, and jars had been \emph{rejected}
by the collectors when they purchased vessels from the huaqueros who
excavated them near Chan Chan.

In his study of collected ceramics from Trujillo, Kroeber established
the outline of a shape typology for Chimú blackware vessels, although it
was not a comprehensive representation of ceramics in use on the north
coast during the periods of Chimú, Inca, and Spanish imperial expansion.
Building on the earlier observations of the high degree of diversity
seen in Chimú blackwares, Kroeber made some important observations that
inform subsequent studies. First, this pottery is widely distributed
along the north coast of Peru, although Kroeber (1930:97) argued for
strong continuity between regions, which would suggest that regional
variations account for little of the overall diversity in the Chimú
assemblage. Second, he noted that blackwares were strongly associated
with mortuary contexts, and such pottery was not commonly seen on the
surface of the urban core of Chan Chan. Such vessels were thus more
indicative of Chimú funerary practices than everyday life. Finally,
Kroeber described how the private collections that local dealers
assembled and sold to museums were already curated in a way that
eliminated certain vessel shapes, so that museum collections could be
considered representative of the original mortuary assemblage. As
Kroeber noted (1930:95) in a later visit to the north coast, one
Trujillo collector ``refused to purchase from the huaqueros most of the
unhandled jars, exceptions being made in favor of effigy pieces, or
occasional plain ones when several vessels were bought in a lot in order
to acquire one or two attractive ones.''

@@

In the years following Kroeber's work, analysis of Chimú ceramics
continued to be oriented around the integration of regional Andean
sequences, and the pottery of the Chimú Empire continued to be discussed
alongside Moche ceramics that were considered to be Early Chimú (e.g.,
Bennett 1937; Willey 1945). It was not until the 1960s that a concerted
effort was made to move beyond the master sequence, to developed a more
fine-grained description of the assemblage associated with the Chimú
Empire as it developed, expanded, and fell under Inca and Spanish
dominion. In 1966, Harry Scheele and Thomas Patterson published a
``preliminary seriation'' of Chimú ceramics. They noted (1966:15) that
one reason that so little work had been done to classify such material
was its abundance: ``probably no other pottery style from the Americas
is so well represented on the shelves of museums and private collectors
throughout the world.''

>>>>

\begin{abstract}
After centuries of looting along Peru's North Coast, archaeologists
acknowledge that the pottery of the Chimú Empire is one of the most
collected Andean artifacts, but also one of the most poorly understood.
Much of the enduring classificatory uncertainty comes from the
problematic provenance of most Chimú vessels, and the fact that the
distinctive blackware identified as Chimú represents the production of
workshops from across an extensive area, during periods of regional
political decentralization (CE 900-1200), imperial growth (CE
1200-1450), and foreign conquest by the Inca (CE 1450-1535) and Spanish
(after CE 1532) empires. This chapter builds on previous seriations and
field observations, using geometric morphometric analysis of a sample of
3D-scanned Chimú bottles from publicly held collections at the
University of Texas at Austin. Since Chimú bottles were formed in
workshops using molds, variations in vessel shape can serve as
indicators of variable practices. We compare a sample of common Chimú
blackware bottles with a sample of \emph{Inca-Chimú} vessels that carry
features typical of the short Inca occupation of the North Coast.
Differences between the two vessel forms offer new lines of evidence
that can guide new interpretations of the imperial history of Peru's
North Coast.
\end{abstract}

\begin{keywords}
Computational Archaeology; Museum Studies; Digital Humanities;
Non-Western Art History; STEM; STEAM
\end{keywords}

@@

The history surrounding the collection of Chimú blackwares helps to
explain why this ubiquitous artifact class remains poorly understood by
professional archaeologists. When viewed as a byproduct of collecting
activities that unfolded over almost a half millennium, the Chimú
ceramic corpus can be seen for what it is: an ambiguous assemblage of
mostly orphaned objects. Almost all Chimú vessels lack a detailed
archaeological provenience, and the chain of ownership for most museum
pieces can rarely be traced back to Peru. Because most ``Chimú'' vessels
cannot be affiliated to a particular context in coastal Peru, their
identification derives from their stylistic attributes, even though some
of the earliest Chimú ceramics documented in archaeological excavations
appeared at the Inca site of Machu Picchu \citep[269]{RN7063} and the
coastal creation shrine of Pachacamac, two sites that lie well beyond
the political control of the Chimú Empire \citep[756]{RN11154}.

This contextual ambiguity intersects with a classificatory problem for
Andean archaeology. Many treat Chimú pottery as typical of the Late
Intermediate Period (CE 900-1476) on the north coast of Peru, a huge and
undifferentiated period during which the Chimú polity established its
capital and local dominion in the Moche Valley, carried out multiple
waves of imperial expansion to the north and south, and fell under the
hegemony of Inca imperial rule (CE 1476-1532). Excavators have noted
that similar blackwares continued to be produced during the half century
of Inca rule, and into the earliest years of the Spanish colonial
period. Without knowing when and where a particular museum object comes
from, archaeologists have faced difficulties in subdividing Chimú
pottery into phases and substyles, which would enable researchers to
study changes and variations within the overall assemblage.

@@

From early on, classifications of Chimú ceramics focused on vessel
attributes. \citet[652-654]{RN11152} noted that coastal tombs had
yielded ``infinite'' formal variations of blackware pots now found in
museum collections. He placed these into five principal categories: (1)
globular vessels with rectangular fields of low-relief decoration; (2)
geometric vessels (cuboid, ellipsoids, ovular); (3) phytomorphic vessels
shaped like squash or fruits; (4) zoomorphic vessels representing a
broad range of fauna; and (5) anthropomorphic vessels. This
classification emphasized aesthetic characteristics of vessel shape,
rather than the kinds of formal and functional differences seen in later
typologies, and Beuchat acknowledged the diversity found within each of
his categories, especially the phytomorphic, zoomorphic, and
anthropomorphic vessels. Without any provenience information to guide
his analysis, Beuchat did not attempt to seriate Chimú pottery or to
distinguish regional variations.

In 1926, Alfred Kroeber made the first concerted effort to describe
different phases of Chimú ceramic production, based on a brief visit to
the Trujillo area that he made as part of a Field Museum expedition to
the central coast. Kroeber noted previous efforts by \citet{RN11154} to
differentiate between early and late ceramics from Trujillo, which could
be tied to the Moche site of Huaca de la Luna/Huaca del Sol and the
Chimú capital of Chan Chan. Kroeber used the term \emph{Late Chimú} to
refer to Trujillo area blackwares that sometimes were found with Inca
ceramics, a designation that encompasses Late Intermediate Period (CE
900-1400s), Inca (CE 1400s-1530s) and some early Colonial pottery. Based
on provenance information that recorded the district where some Late
Chimú vessels originated, \citet[11]{RN11151} identified a broad
distribution that stretched from Piura to Casma, with some vessels found
as far south as the Nasca drainage. To build a typology and sequence of
Late Chimú vessels, he studied the pottery from several collections that
he purchased in Trujillo on behalf of the Field Museum.

Kroeber's shape typology for Late Chimú pottery discusses vessels that
were common in Moche and Chimú assemblages, and he lists
\citep[22]{RN11151} about a dozen Late Chimú shapes, including
stirrup-mouth vessels, double jars, double-spout vessels, head-and-spout
vessels, figure-and-spout vessels, globular bowls, lipped pots (with and
without handles), jars with flaring mouth, flat-handled jars, and jars
with tapering spouts. The list of shapes also includes Inca narrow-mouth
jars (\emph{aryballus}) and cups (\emph{kero}) in the Late Chimú
assemblage. In his decorative analysis of vessels in private
collections, \citet[27-28]{RN11151} identified the \emph{face vase} and
\emph{rotund figure jar} as additional shapes found in the assemblage.
The discussion of Late Chimú blackware recorded the relative abundance
of the most common shapes found in two private collections that he saw
while visiting Trujillo. The Mansiche and Jacoms collections comprised
more than 200 vessels, but Kroeber estimated that hundreds of
utilitarian pots, plates, and jars had been \emph{rejected} by the
collectors when they purchased vessels from the huaqueros who excavated
them near Chan Chan.

In his study of collected ceramics from Trujillo, Kroeber established
the outline of a shape typology for Chimú blackware vessels, although it
was not a comprehensive representation of ceramics in use on the north
coast during the periods of Chimú, Inca, and Spanish imperial expansion.
Building on the earlier observations of the high degree of diversity
seen in Chimú blackwares, Kroeber made some important observations that
inform subsequent studies. First, this pottery is widely distributed
along the north coast of Peru, although \citet[97]{RN11155} argued for
strong continuity between regions, which would suggest that regional
variations account for little of the overall diversity in the Chimú
assemblage. Second, he noted that blackwares were strongly associated
with mortuary contexts, and such pottery was not commonly seen on the
surface of the urban core of Chan Chan. Such vessels were thus more
indicative of Chimú funerary practices than everyday life. Finally,
Kroeber described how the private collections that local dealers
assembled and sold to museums were already curated in a way that
eliminated certain vessel shapes, so that museum collections could be
considered representative of the original mortuary assemblage. As
Kroeber noted \citep[95]{RN11155} in a later visit to the north coast,
one Trujillo collector ``refused to purchase from the huaqueros most of
the unhandled jars, exceptions being made in favor of effigy pieces, or
occasional plain ones when several vessels were bought in a lot in order
to acquire one or two attractive ones.''

@@

In the years following Kroeber's work, analysis of Chimú ceramics
continued to be oriented around the integration of regional Andean
sequences, and the pottery of the Chimú Empire continued to be discussed
alongside Moche ceramics that were considered to be Early Chimú
\citep{RN11156,RN11157}. It was not until the 1960s that a concerted
effort was made to move beyond the master sequence, to developed a more
fine-grained description of the assemblage associated with the Chimú
Empire as it developed, expanded, and fell under Inca and Spanish
dominion. In 1966, Harry Scheele and Thomas Patterson published a
``preliminary seriation'' of Chimú ceramics. They noted
\citep[15]{RN11158} that one reason that so little work had been done to
classify such material was its abundance: ``probably no other pottery
style from the Americas is so well represented on the shelves of museums
and private collectors throughout the world.''
